---
title: "Using RcppArmadillo to price European Put Options"
author: "Davis Vaughan"
date: "2/27/2018"
output: html_document
---

## Introduction

In the quest for ever faster code, one generally begins exploring ways to integrate
C++ with R using `Rcpp`. This post provides an example of multiple implementations
of a European Put Option pricer. The implementations are done in pure R, pure 
`Rcpp` using some `Rcpp` sugar functions, and then in `Rcpp` using `RcppArmadillo`,
which exposes the incredibly powerful linear algebra library, 
[`Armadillo`](http://arma.sourceforge.net/).

The value of a European Put Option has the closed form solution:

$$ V = K e^{-rt} N(-d_2) - S e^{-yt} N(-d_1) $$

Where:

$$ 
\begin{equation}
  \begin{aligned}
V    &= \text{Value of the option} \\
r    &= \text{Risk free rate} \\
y    &= \text{Dividend yield} \\
t    &= \text{Time to expiry} \\
S    &= \text{Current stock price} \\
K    &= \text{Strike price} \\
N(.) &= \text{Normal CDF}
  \end{aligned}
\end{equation}
$$
And:

$$ 
\begin{equation}
  \begin{aligned}
d_1    &= \frac{log(\frac{S}{K}) + (r - y + \frac{1}{2} \sigma^2)t}{\sigma \sqrt{t}}  \\
d_2    &= d_1 - \sigma \sqrt{t}\\
  \end{aligned}
\end{equation}
$$

Armed with the formulas, we can create the pricer using just R.

```{r}
put_option_pricer <- function(s, k, r, y, t, sigma) {

  d1 <- (log(s / k) + (r - y + sigma^2 / 2) * t) / (sigma * sqrt(t))
  d2 <- d1 - sigma * sqrt(t)

  V <- pnorm(-d2) * k * exp(-r * t) - s * exp(-y * t) * pnorm(-d1)

  V
}

# Valuation with 1 stock price
put_option_pricer(s = 55, 60, .01, .02, 1, .05)

# Valuation across multiple prices
put_option_pricer(s = 55:60, 60, .01, .02, 1, .05)
```

Let's see what we can do with `Rcpp`. Besides explicitely stating the types of the
variables, not much has to change. We can even use the sugar function, `Rcpp::pnorm()`,
to keep the syntax as close to R as possible. We don't have to use `Rcpp::` here to 
qualify our `log()` and `pnorm()` calls, but I find it to be good practice.

```{r}
library(Rcpp)

cppFunction('
  NumericVector put_option_pricer_rcpp(NumericVector s, double k, double r, double y, double t, double sigma) {

    NumericVector d1 = (Rcpp::log(s / k) + (r - y + sigma * sigma / 2.0) * t) / (sigma * sqrt(t));
    NumericVector d2 = d1 - sigma * sqrt(t);
    
    NumericVector V = Rcpp::pnorm(-d2) * k * exp(-r * t) - s * exp(-y * t) * Rcpp::pnorm(-d1);
    return V;
}')

# Valuation with 1 stock price
put_option_pricer_rcpp(s = 55, 60, .01, .02, 1, .05)

# Valuation across multiple prices
put_option_pricer_rcpp(s = 55:60, 60, .01, .02, 1, .05)
```

Finally, let's look at `RcppArmadillo`. Armadillo has a number of object types,
including `mat`, `colvec`, and `rowvec`. Here, we just use `colvec` to represent
a column vector of prices. By default in Armadillo, `*` represents matrix multiplication, and
`%` is used for element wise multiplication. We need to make this change to 
element wise multiplication in 1 place, but otherwise the changes are just
switching out the types and the sugar functions for Armadillo specific functions.

Note that the `arma::normcdf()` function is in the development version of `RcppArmadillo`,
which is `0.8.400.0.0` at the time of writing and additionally requires `C++11` as a plugin.

```{r}
library(RcppArmadillo)

cppFunction('
  arma::colvec put_option_pricer_arma(arma::colvec s, double k, double r, double y, double t, double sigma) {
  
    arma::colvec d1 = (arma::log(s / k) + (r - y + sigma * sigma / 2.0) * t) / (sigma * sqrt(t));
    arma::colvec d2 = d1 - sigma * sqrt(t);
    
    // Notice the use of % to represent element wise multiplication
    arma::colvec V = arma::normcdf(-d2) * k * exp(-r * t) - s * exp(-y * t) % arma::normcdf(-d1); 

    return V;
  }', 
  depends = "RcppArmadillo",
  plugins = "cpp11")

# Valuation with 1 stock price
put_option_pricer_arma(s = 55, 60, .01, .02, 1, .05)

# Valuation across multiple prices
put_option_pricer_arma(s = 55:60, 60, .01, .02, 1, .05)
```

Finally, we can run a speed test to see which comes out on top.

```{r}
s <- matrix(seq(0, 100, by = .0001), ncol = 1)

rbenchmark::benchmark(
  put_option_pricer(s, 60, .01, .02, 1, .05),
  put_option_pricer_arma(s, 60, .01, .02, 1, .05),
  put_option_pricer_rcpp(s, 60, .01, .02, 1, .05), 
  order = "relative", 
  replications = 100
)[,1:4]
```

Interestingly, `Rcpp` comes out on top with both C++ implementations being faster than pure R,
but this isn't the most computationally intensive problem so the differences
aren't too extreme. The nicest thing about all of this is that it shows off the 
"embarassment of riches" that we have in the R and C++ ecosystem for multiple ways
of solving the same problem.